---
title: 'Bonanza Creek Snowshoe Hares: Exploratory Report'
author: "Anastasia Kunz"
date: "11/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(effsize)
library(broom)
library(palmerpenguins)
library(ggbeeswarm)
library(kableExtra)
library(here)
library(lubridate)
```

# Bonanza Creek Snowshoe Hares: Exploratory Report
 Anastasia Kunz
 
```{r}
# Read in data
hare_data <- read_csv(here("data", "bonanza_hares.csv"))
```
 
## 1. Overview
Just cursory overview of what is included DO THIS LAST BUT DONT FORGET!!!!


## 2. Data and Analysis
A 3 - 4 sentence section in which you briefly describe the data, including citation & credit as needed, and provide an overview of the main analysis, statistical methods and tools used.
....All analyses are in R version 4.0.2 using RStudio version 1.3.1073.



Possible map and figure?? maybe

## 3. Exploratory Findings
### A. Annual juvenile hare trap counts

```{r}
#filter for juvenile hares and create year out of date column
juvenile_hare_yr <- hare_data %>% 
  filter(age == "j") %>% 
  mutate(date_converted = mdy(date)) %>% 
  mutate(year = year(date_converted))%>% 
  count(year)

# create visualization
ggplot(juvenile_hare_yr, aes(x = year, y = n)) +
  geom_col(size = 2, fill = "dodgerblue2") +
  theme_bw()+
  labs(
    x = "Year",
    y = "Juvenile Hare Count",
    title = "Total Juvenile Hares Per Year",
    subtitle = "1998 - 2012"
  )
  
```

**Figure 1.** Total juvenile hares recorded per year at all sites. 

In 1999 we see the most juvenile hares trapped with a maximum count of 126 juvenile hares. However in two years, 2002 and 2009, no juvenile hares were counted. It is important to note the consistency of days each year that hares were recorded. In some years there are significantly more days in which data is recorded than in other years. In addition, age of the hares is inconsistently reported, so it is difficult to comment on population trends without age assessment of the hares that were trapped.

### B. Visualize Juvenille Hare Weights

```{r}
#juvenile hare weight by sex and site 
juvenile_hare_weight <- hare_data %>% 
  filter(age == "j")

grid.labs<- c("Bonanza Black Spruce", "Bonanza Mature", "Bonanza Riparian")
names(grid.labs) <- c("bonbs", "bonmat", "bonrip")

ggplot(juvenile_hare_weight, aes(x = sex, y = weight))+
  geom_boxplot(aes(fill = sex))+
  facet_wrap(~ grid, labeller = labeller(grid = grid.labs))+
  theme_bw()+
  labs(
    x = "Sex",
    y = "Weight (g)",
    title = "Juvenile Weight by Sex at Each Site",
    subtitle = "1998 - 2012"
  )+
  scale_x_discrete(labels=c("f" = "Female", "m" = "Male",
                              "NA" = "NA"))+
  theme(legend.position = "none")
```

**Figure 2.** Box and whisker plot showing female, male and NA weight. FIX THIS CAPTION


### C. Juvenile Weight Comparison (Male & Female)

**Table 1.**  CAPTION HERE LOOK UP HOW TO GE
```{r}
#compare mean weights for juvenile male and female hares in the study: table with mean stdev and sample size 
mf_hare_weight<- juvenile_hare_weight %>% 
  filter(sex == "f" | sex == "m") %>% 
  mutate(sex_new = case_when(
    sex == "m" ~ "Male",
    sex == "f" ~ "Female"
  )) %>% 
  group_by(sex_new) %>% 
  summarize(mean_g = mean(weight, na.rm = TRUE),
            standard_deviation = sd(weight, na.rm = TRUE),
            sample_size = n())
mf_hare_weight %>% 
  kable(col.names = c("Sex", "Mean Weight (g)", "Standard Deviation (g)", "Sample Size")) %>% 
  kable_styling(full_width = FALSE)
  
  
```



```{r}
# T-Test Code

m_hare <- juvenile_hare_weight %>% 
  filter(sex == "m") %>% 
  pull(weight)

f_hare_na <-juvenile_hare_weight %>% 
  filter(sex == "f") %>% 
  pull(weight)

f_hare <- na.omit(f_hare_na)

```


```{r, include = FALSE}
#Exploratory Graphs
hist(m_hare) #looks pretty normal
hist(f_hare) #looks pretty normal too
qqnorm(m_hare) #bit of a curve, but overall pretty normal too
qqnorm(f_hare) #normal

# Main Takeaways: Generally normal, no glaring issues. The male hare weight distribution is perhaps a little off from a normal distribution, but a t-test should suffice for mean weight comparison.

hare_ttest <- t.test(m_hare, f_hare)

# Get the tidy model results to call outputs in-line: 
hare_ttest_tidy <- tidy(hare_ttest)

# hare_ttest_tidy

# Find  means and std dev 
m_weight_mean <- mean(m_hare)
f_weight_mean <- mean(f_hare)
m_weight_sd <- sd(m_hare)
f_weight_sd <- sd(f_hare)

# Cohen's D Effect Size
hare_effsize <- cohen.d(m_hare, f_hare)

```
NEED COHENS d EFFECT SIZE TWO SAMPLE T TEST AND DIFFERENCE IN MEANS 2-3 sentences HERE 
LOOK UP HOW TO REFERENCE MODEL IN REPORT 

### D. Relationship Between Juvenile Weight and Hind Foot Length

```{r}
ggplot(juvenile_hare_weight, aes(x = hindft, y = weight))+
  geom_point()+
  geom_smooth(method = lm, color = "dodgerblue2")+
  theme_bw()+
  labs(
    x = "Hindfoot Length (mm)",
    y = "Weight (g)",
    title = "Relationship of Snowshoe Hare Weight and Hindfoot Length",
    subtitle = "1998 - 2012"
  )
```

**Figure 3.** CAPTION HERE

outcome of linear regression, sllope interpretation, R2 and pearsons r, what assumptions of linear regression may be a concern. how well or not well linear reg describes relationship. Worries?


## 4. Summary

Overarching summ sentence
  - bullet pts are good here

## 5. Citations

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14











